Convergence rate of MAP estimates \\ for the exponential family

### 2021-09-10 main.tex
Add discussion of the entropic barrier paper

As for the MAP, we did not manage to get an asymptotically tight upper bound. Nevertheless, using inequality~\eqref{eq:log_bound}, we did get an interesting upper bound. 
To start, let us recall that the MAP of $\cN(0,\mu_*)$ is $\hat \mu_n = \frac{n_0 \mu_0 + \sum_i X_i^2}{n_0 + n}$.
Its expectation is simply $\mu_n= \frac{n_0 \mu_0 + n \mu^*}{n_0 + n}$.
We now present a lemma on its expected inverse.
\begin{lemma}[Expected MAP Natural Parameter]
	For any $n\geq 2$, the expectation of the natural parameter of the MAP of $\cN(0,\mu_*)$ is bounded as
	\begin{align}
		\frac{\mu^*}{\mu_n}
		\leq \expect{\frac{\mu^*}{\MAPm}} 
		= \expect{\frac{\MAPt}{\natp^*}} 
		\leq \frac{1}{\frac{\mu_n}{\mu^*} - \frac{2}{n+n_0}} \; .
	\label{eq:theta_expectation_bound}
	\end{align}
\end{lemma}
Note that the lower bound is a simple consequence of the convexity of the inverse function. 
Using this lemma, along with the log upper bound on the logarithm~\eqref{eq:log_bound}, we get a convergence rate for the MAP. 
\begin{example}
\begin{theorem}[MAP Bound]
 Introducing $a=n_0 \frac {\mu_0 }{ \mu^* }$, the expected suboptimality of the MAP of $\cN(0,\mu^*)$ with prior hyper-parameters $(n_0,\mu_0)$ is
 \begin{equation}
	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MAP})}
	\leq \begin{cases}
		\bregmanconj( \mu_*; \mu_0) \ \text{ if } \ n=0, \\
		\bregmanconj( \mu_*; \mu_0) + \inv{a} \ \text{if}\ n=1,\\
		\frac{1}{a+n-2} + \frac{\half(1 - \frac{\mu_0}{\mu^*})^2}{(\frac{\mu_0}{\mu^*}+\frac{n-2}{n_0})(1 + \frac{n}{n_0} )} \ \text{if}\  n\geq 2 \ \text{and}\ a=n_0 \frac{\mu_0}{\mu^*} >0
	\end{cases}
	\label{eq:MAP_rate}
\end{equation}
\end{theorem}
\end{example}
Using the same bound on the log~\eqref{eq:log_bound}, we  get a loose bound on the MLE, which is useful for comparison purposes.
We relate these loose convergence rates of MAP and MLE in Figure~\ref{fig:MAP_rate}.
\begin{example}
\begin{corollary}[MLE Loose Bound]
\begin{equation}
	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE}) }
	\leq \inv{n-2} \; .
	\label{eq:MLE_loose_rate}
\end{equation}
\end{corollary}
\end{example}

@@

\paragraph{Stochastic Setting.}
This proof does not transpose immediately to the stochastic setting.
First off, we do not have a decreasing sequence anymore
\begin{align}
	f(\natp_{t+1}, X_t)  \leq f(\natp_{t}, X_t)  
	- \inv{\lr_t}\left (\cB(\natp_{t+1} ; \natp_{t}) + \cB(\natp_{t} ; \natp_{t+1}) \right) \\
	\implies \expect{f(\natp_{t+1}, X_t)} \leq f(\natp_{t}) 
	- \inv{\lr_t}\expect{\cB(\natp_{t+1} ; \natp_{t}) + \cB(\natp_{t} ; \natp_{t+1}))}  \; .
\end{align}
Indeed the inequality bears on $\expect{f(\natp_{t+1}, X_t)}$, which features a correlation between $\natp_{t+1}$ and $X_t$, rather than $f(\natp_{t+1})$.
Similarly, applying the three points lemma with $\natp^*$, taking the tower expectation, and telescoping the sum, we get
\begin{align}
	\sum_{t=0}^{t-1} \lr_t \expect{f(\natp_{t+1},X_t) - f(\natp^*)}
	\leq \cB( \natp^*; \natp_{0})  \; .
\end{align}
To pursue further, we need an assumption on the quality of the stochastic estimates $f(. ; x)$. 
The most straightforward such assumption is 
\begin{align}
\expect{f(\natp_{t+1}) - f(\natp_{t+1} ; X_{t})} \leq \lr_t \sigma^2 \; ,
\label{eq:proximal_variance}
\end{align}
e.g. we relate the stochastic function value on the next iterate to the function value.
Then, taking the iterate with the least expected loss, we get the rate
\begin{align}
	 \min_{1\leq t \leq T} \expect{ f(\natp_t)} - f(\natp^*)
	\leq \frac{\cB( \natp^*; \natp_{0}) }{\sum_{t=0}^{T-1} \lr_t}
	+ \frac{\sum_{t=0}^{T-1} \lr_t^2  }{\sum_{t=0}^{T-1} \lr_t} \sigma^2 \; .
\end{align}
which is very analogous to standard rates for SGD.
It does converge, whenever $\sum_{t=0}^{T-1} \lr_t$ diverges to $+\infty$ much faster than $\sum_{t=0}^{T-1} \lr_t^2$.
When $\lr_t \propto t^{-\half}$ for instance, we get a rate $O(\frac{\log(t)}{t^{\half[3]}})$.
But in our case, $\lr_t \propto t^{-1}$, and we merely get a $O(\inv{\log(t)})$ rate.

@@

\paragraph{Stochastic Mirror Descent}
Let $f(\natp, x) = -\log p(x|\natp)$.
MAP iterations can be cast as stochastic mirror descent (SMD),  with $g_t = \nabla f(\natp_t, x_{t+1}) = \logpart(\natp_t) - T(x_{t+1})$ as a stochastic estimate  of $\nabla f(\natp_t)$.
Resuming from the Stochastic Bregman Proximal Point formula~\eqref{eq:SBPP} and manipulating the Bregman yields
\begin{align}
    \hat \natp_{n+1} 
    &= \argmin_\natp 
    - \langle T(x_{t+1}), \natp \rangle  + \logpart(\natp) + (n_0 + n) \bregman(\natp ; \natp_n) \\
    %&(n_0+n)\left(\logpart(\natp) - \logpart(\natp_n) - \langle \nabla \logpart(\natp_n)  , \natp - \natp_n \rangle \right)  \\
    &= \argmin_\natp - \langle T(x_{t+1}), \natp \rangle   + \logpart(\natp_n) + \langle \nabla \logpart(\natp_n)  , \natp - \natp_n \rangle + (n_0+n + 1) \bregman(\natp ; \natp_n) \\
    &= \argmin_\natp \linear_f(\natp;\natp_n, x_{t+1}) + (n_0 + n + 1) \bregman(\natp ; \natp_n)
\end{align}
where $\linear_f(\natp;\natp_n, x_{t+1})$ is the stochastic linearization of $f$ at $\natp_n$ evaluated at $\natp$ with randomness coming from $x_{t+1}$. This is the formula for stochastic mirror descent (SMD) applied to $f$ with mirror map $\logpart$ and step-size $\gamma_n = \inv{n_0 + n + 1}$.

@@

@@

\citet{ostrovskii2021finite}  shows a convergence rate for the MLE (or other $M$-estimators)   when  the log-likelihood is self-concordant. This convergence rate holds only after a given number of samples have been seen. This is because they use the fact that self-concordant losses are upper bounded by a quadratic in a neighborhood of the optimum. 
\TODO{Sanity check literature search with Bach's references.}

>>>>

As for the MAP, we did not manage to get an asymptotically tight upper bound. Nevertheless, using inequality~\eqref{eq:log_bound}, we did get an interesting upper bound. 
To start, let us recall that the MAP of $\cN(0,\mu_*)$ is $\hat \mu_n = \frac{n_0 \mu_0 + \sum_i X_i^2}{n_0 + n}$.
Its expectation is simply $\mu_n= \frac{n_0 \mu_0 + n \mu^*}{n_0 + n}$.
We now present a lemma on its expected inverse.
\begin{lemma}[Expected MAP Natural Parameter]
	Let's define $a=n_0 \frac {\mu_0 }{ \mu^* }$. For any $n\geq 2$, the expectation of the natural parameter of the MAP of $\cN(0,\mu_*)$ is bounded as
	\begin{align}
		\frac{\mu^*}{\mu_n} = \frac{n_0+n}{a+n}
		\leq \expect{\frac{\mu^*}{\MAPm}} 
		= \expect{\frac{\MAPt}{\natp^*}} 
		\leq \frac{n_0+n}{a+n-2} \; .
	\label{eq:theta_expectation_bound}
	\end{align}
\end{lemma}
Note that the lower bound is a simple consequence of the convexity of the inverse function. 
Using this lemma, along with the log upper bound on the logarithm~\eqref{eq:log_bound}, we get a convergence rate for the MAP. 
\begin{example}
\begin{theorem}[MAP Bound]
 Defining $a=n_0 \frac {\mu_0 }{ \mu^* }$, the expected suboptimality of the MAP of $\cN(0,\mu^*)$ with prior hyper-parameters $(n_0,\mu_0)$ is
 \begin{equation}
	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MAP})}
	\leq \begin{cases}
		\bregmanconj( \mu_*; \mu_0) \ \text{ if } \ n=0, \\
		\bregmanconj( \mu_*; \mu_0) + \inv{a} \ \text{if}\ n=1,\\
		\frac{1}{a+n-2} + \frac{\half(1 - \frac{\mu_0}{\mu^*})^2}{\frac{a+n-2}{n_0}(1 + \frac{n}{n_0} )} \ \text{if}\  n\geq 2 \ \text{and}\ a=n_0 \frac{\mu_0}{\mu^*} >0
	\end{cases}
	\label{eq:MAP_rate}
\end{equation}
\end{theorem}
\end{example}
Using the same bound on the log~\eqref{eq:log_bound}, we  get a loose bound on the MLE, which is useful for comparison purposes.
We relate these loose convergence rates of MAP and MLE in Figure~\ref{fig:MAP_rate}.
\begin{example}
\begin{corollary}[MLE Loose Bound]
\begin{equation}
	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE}) }
	\leq \inv{n-2} \; .
	\label{eq:MLE_loose_rate}
\end{equation}
\end{corollary}
\end{example}

@@

\paragraph{Stochastic Setting.}
This proof does not transpose immediately to the stochastic setting.
First off, we do not have a decreasing sequence anymore
\begin{align}
	f(\natp_{t+1}, X_t)  \leq f(\natp_{t}, X_t)  
	- \inv{\lr_t}\left (\cB(\natp_{t+1} ; \natp_{t}) + \cB(\natp_{t} ; \natp_{t+1}) \right) \\
	\implies \expect{f(\natp_{t+1}, X_t)} \leq f(\natp_{t}) 
	- \inv{\lr_t}\expect{\cB(\natp_{t+1} ; \natp_{t}) + \cB(\natp_{t} ; \natp_{t+1}))}  \; .
\end{align}
Indeed the inequality bears on $\expect{f(\natp_{t+1}, X_t)}$, which features a correlation between $\natp_{t+1}$ and $X_t$, rather than $f(\natp_{t+1})$.
Similarly, applying the three points lemma with $\natp^*$, taking the tower expectation, and telescoping the sum, we get
\begin{align}
	\sum_{t=0}^{t-1} \lr_t \expect{f(\natp_{t+1},X_t) - f(\natp^*)}
	\leq \cB( \natp^*; \natp_{0})  \; .
\end{align}
To pursue further, we need an assumption on the quality of the stochastic estimates $f(. ; x)$. 
The most straightforward such assumption is 
\begin{align}
\expect{f(\natp_{t+1}) - f(\natp_{t+1} ; X_{t})} \leq \lr_t \sigma^2 \; ,
\label{eq:proximal_variance}
\end{align}
e.g. we relate the stochastic function value on the next iterate to the function value.
Then, taking the iterate with the least expected loss, we get the rate
\begin{align}
	 \min_{1\leq t \leq T} \expect{ f(\natp_t)} - f(\natp^*)
	\leq \frac{\cB( \natp^*; \natp_{0}) }{\sum_{t=0}^{T-1} \lr_t}
	+ \frac{\sum_{t=0}^{T-1} \lr_t^2  }{\sum_{t=0}^{T-1} \lr_t} \sigma^2 \; .
\end{align}
which is very analogous to standard rates for SGD.
It does converge, whenever $\sum_{t=0}^{T-1} \lr_t$ diverges to $+\infty$ much faster than $\sum_{t=0}^{T-1} \lr_t^2$.
When $\lr_t \propto t^{-\half}$ for instance, we get a rate $O(\frac{\log(t)}{\sqrt{t}})$.
But in our case, $\lr_t \propto t^{-1}$, and we merely get a $O(\inv{\log(t)})$ rate.

@@

\paragraph{Stochastic Mirror Descent}
Let $f(\natp, x) = -\log p(x|\natp)$.
MAP iterations can be cast as stochastic mirror descent (SMD),  with $g_t = \nabla f(\natp_t, x_{t+1}) = \logpart(\natp_t) - T(x_{t+1})$ as a stochastic estimate  of $\nabla f(\natp_t)$.
Resuming from the Stochastic Bregman Proximal Point formula~\eqref{eq:SBPP}, we write $\MAPt$ for the n-th MAP iterate, and do some Bregman manipulations
\begin{align}
    \hat \natp_{n+1}
    &= \argmin_\natp 
    - \langle T(x_{t+1}), \natp \rangle  + \logpart(\natp) + (n_0 + n) \bregman(\natp ; \natp_n) \\
    %&(n_0+n)\left(\logpart(\natp) - \logpart(\natp_n) - \langle \nabla \logpart(\natp_n)  , \natp - \natp_n \rangle \right)  \\
    &= \argmin_\natp - \langle T(x_{t+1}), \natp \rangle   + \logpart(\natp_n) + \langle \nabla \logpart(\natp_n)  , \natp - \natp_n \rangle + (n_0+n + 1) \bregman(\natp ; \natp_n) \\
    &= \argmin_\natp \linear_f(\natp;\natp_n, x_{t+1}) + (n_0 + n + 1) \bregman(\natp ; \natp_n)
\end{align}
where $\linear_f(\natp;\natp_n, x_{t+1})$ is the stochastic linearization of $f$ at $\natp_n$ evaluated at $\natp$ with randomness coming from $x_{t+1}$. This is the formula for stochastic mirror descent (SMD) applied to $f$ with mirror map $\logpart$ and step-size $\gamma_n = \inv{n_0 + n + 1}$.

@@

\paragraph{Questions.}
Is there already an SC analysis of proximal point methods. (that’s not affine invariant)?
If the log-partition is SC, then what does it mean for its gradient, the mapping from natural to mean parameters ?

@@

\citet{bubeck2015entropic} shows that the entropy of an exponential family defined on a compact support $\cX$ is a self-concordant barrier on this compact set. They extend the definition of self-concordance to $\nu$-self-concordance, stating that the gradient should be bounded by the square root of the hessian along any direction. In 1 dimension, their result also means that the log-partition is self-concordant as well. In general, does this mean that the log-partition is self-concordant  (e.g. is there a link between the dual function on a line and the line of the primal , probably not)?

In the realm of self-concordance, \citet{ostrovskii2021finite}  shows a finite sample convergence rate for the MLE (or other $M$-estimators)   when  the log-likelihood is self-concordant. This convergence rate holds only after a given number of samples have been seen, and they give an innteresting formula for this number. To get this results, they use the fact that self-concordant losses are upper bounded by a quadratic in a neighborhood of the optimum.
